# Base subsetting  {#basicR}

teaching: 50  
exercises: 30  
adapted from: https://datacarpentry.org/r-socialsci/01-intro-to-r/index.html  

questions:  

- How can subsets be extracted from vectors and data frames?  
- How does R treat missing values?  
- How can we deal with missing values in R?  

objectives:  

- Subset and extract values from vectors.  
- Analyze vectors with missing data.  

keypoints:  

- Access individual values by location using `[]`.  
- Access arbitrary sets of data using `[c(...)]`.  
- Use logical operations and logical vectors to access subsets of data. 

## Tidyverse and base R

As a programming language, R provides a lot of flexibility. Where other 
programming languages (e.g., Python) require a very particular syntax, and there
is usually just one right way to do something, R developers can use whatever 
syntax they would like. This has led to a proliferation of R syntaxes, or many 
ways to "say" the same thing. A [cheatsheet](https://github.com/rstudio/cheatsheets/raw/master/syntax.pdf) shows
how to do the same tasks many ways in R.

As you develop as an R user and programmer, you will learn to mix and match 
these approaches in your work. Usually, there is one way that is easier (at least
for a particular person) and one that is more challenging. So far, we have 
focused on Tidyverse approaches, but we also want you to see the more general
base R approach to some tasks. 

In particular, we want to come back to subsetting vectors and data frames. When
we discussed `dplyr`, we used the `filter` command to retrieve certain rows from a
dataframe, and the `select` command to retrieve certain columns. We are going to
move to a slightly more matrix-oriented approach, using the square brackets 
`[ , ]`.


## Subsetting vectors

If we want to extract one or several values from a vector, we must provide one 
or several indices in square brackets. For instance:

```{r, results='show', purl=FALSE}
respondent_wall_type <- c("muddaub", "burntbricks", "sunbricks")
respondent_wall_type[2]
respondent_wall_type[c(3, 2)]
```

We can also repeat the indices to create an object with more elements than the 
original one:

```{r, results='show', purl=FALSE}
more_respondent_wall_type <- respondent_wall_type[c(1, 2, 3, 2, 1, 3)]
more_respondent_wall_type
```

R indices start at 1. Programming languages like Fortran, MATLAB, Julia, and R 
start counting at 1, because that's what human beings typically do. Languages in 
the C family (including C++, Java, Perl, and Python) count from 0 because that's
simpler for computers to do.

### Conditional subsetting

Another common way of subsetting is by using a logical vector. `TRUE` will 
select the element with the same index, while `FALSE` will not:

```{r, results='show', purl=FALSE}
hh_members <- c(3, 7, 10, 6)
hh_members[c(TRUE, FALSE, TRUE, TRUE)]
```

Typically, these logical vectors are not typed by hand, but are the output of 
other functions or logical tests. For instance, if you wanted to select only 
the values above 5:

```{r, results='show', purl=FALSE}
hh_members > 5    # will return logicals with TRUE for the indices that meet the condition
## so we can use this to select only the values above 5
hh_members[hh_members > 5]
```

You can combine multiple tests using `&` (both conditions are true, AND) or `|` 
(at least one of the conditions is true, OR):

```{r, results='show', purl=FALSE}
hh_members[hh_members < 3 | hh_members > 5]
hh_members[hh_members >= 7 & hh_members == 3]
```

Here, `<` stands for "less than", `>` for "greater than", `>=` for "greater than
or equal to", and `==` for "equal to". The double equal sign `==` is a test for 
numerical equality between the left and right hand sides, and should not be
confused with the single `=` sign, which performs variable assignment (similar to `<-`).

A common task is to search for certain strings in a vector.  One could use the 
"or" operator `|` to test for equality to multiple values, but this can quickly 
become tedious. The function `%in%` allows you to test if any of the elements of
a search vector are found:

```{r, results='show', purl=FALSE}
possessions <- c("car", "bicycle", "radio", "television", "mobile_phone")
possessions[possessions == "car" | possessions == "bicycle"] # returns both car and bicycle
possessions %in% c("car", "bicycle", "motorcycle", "truck", "boat")
possessions[possessions %in% c("car", "bicycle", "motorcycle", "truck", "boat")]
```

## Missing data

As R was designed to analyze datasets, it includes the concept of missing data 
(which is uncommon in other programming languages). Missing data are represented
in vectors as `NA`.

When doing operations on numbers, most functions will return `NA` if the data 
you are working with include missing values. This feature makes it harder to 
overlook the cases where you are dealing with missing data. You can add the 
argument `na.rm=TRUE` to calculate the result while ignoring
the missing values.

```{r, purl=FALSE}
rooms <- c(2, 1, 1, NA, 4)
mean(rooms)
max(rooms)
mean(rooms, na.rm = TRUE)
max(rooms, na.rm = TRUE)
```

If your data include missing values, you may want to become familiar with the 
functions `is.na()`, `na.omit()`, and `complete.cases()`. See below for examples.


```{r, purl=FALSE}
## Extract those elements which are not missing values.
rooms[!is.na(rooms)]

## Returns the object with incomplete cases removed. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
na.omit(rooms)

## Extract those elements which are complete cases. The returned object is an atomic vector of type `"numeric"` (or `"double"`).
rooms[complete.cases(rooms)]
```

Recall that you can use the `typeof()` function to find the type of your atomic vector.

> ## Exercise
>
> 1. Using this vector of rooms, create a new vector with the NAs removed.
>
>     ```r
>     rooms <- c(1, 2, 1, 1, NA, 3, 1, 3, 2, 1, 1, 8, 3, 1, NA, 1)
>     ```
> 2. Use the function `median()` to calculate the median of the `rooms` vector.
>
> 3. Use R to figure out how many households in the set use more than 2 rooms for sleeping.
>
> > ## Solution
> > ```{r, answer=TRUE}
> > rooms <- c(1, 2, 1, 1, NA, 3, 1, 3, 2, 1, 1, 8, 3, 1, NA, 1)
> > rooms_no_na <- rooms[!is.na(rooms)]
> > # or
> > rooms_no_na <- na.omit(rooms)
> > # 2.
> > median(rooms, na.rm = TRUE)
> > # 3.
> > rooms_above_2 <- rooms_no_na[rooms_no_na > 2]
> > length(rooms_above_2)
> > ```



## Indexing and subsetting data frames

We have seen how to use square brackets to index vectors. We can extend the same
concept to dataframes. 

Consider our interviews data frame. It has rows and columns, so it has 2 
dimensions. If we want to
extract some specific data from it, we need to specify the "coordinates" we
want from it. Row numbers come first, followed by column numbers. However, note
that different ways of specifying these coordinates lead to results with
different classes.


```{r, purl=FALSE}
## first element in the first column of the data frame (as a vector)
interviews[1, 1]
## first element in the 6th column (as a vector)
interviews[1, 6]
## first column of the data frame (as a vector)
interviews[[1]]
## first column of the data frame (as a data.frame)
interviews[1]
## first three elements in the 7th column (as a vector)
interviews[1:3, 7]
## the 3rd row of the data frame (as a data.frame)
interviews[3, ]
## equivalent to head_interviews <- head(interviews)
head_interviews <- interviews[1:6, ]
```

`:` is a special function that creates numeric vectors of integers in increasing
or decreasing order, test `1:10` and `10:1` for instance.

You can also exclude certain indices of a data frame using the "`-`" sign:

```{r, purl=FALSE}
interviews[, -1]          # The whole data frame, except the first column
interviews[-c(7:131), ]   # Equivalent to head(interviews)
```

Data frames can be subset by calling indices (as shown previously), but also by calling their column names directly:

```{r, eval = FALSE, purl=FALSE}
interviews["village"]       # Result is a data frame
interviews[, "village"]     # Result is a data frame
interviews[["village"]]     # Result is a vector
interviews$village          # Result is a vector
```

In RStudio, you can use the autocompletion feature to get the full and correct names of the columns.

> ## Exercise
>
> 1. Create a data frame (`interviews_100`) containing only the data in
>    row 100 of the `interviews` dataset.
>
> 2. Notice how `nrow()` gave you the number of rows in a data frame?
>
>      * Use that number to pull out just that last row in the data frame.
>      * Compare that with what you see as the last row using `tail()` to make
>        sure it's meeting expectations.
>      * Pull out that last row using `nrow()` instead of the row number.
>      * Create a new data frame (`interviews_last`) from that last row.
>
> 3. Use `nrow()` to extract the row that is in the middle of the data frame.
>    Store the content of this row in an object named `interviews_middle`.
>
> 4. Combine `nrow()` with the `-` notation above to reproduce the behavior of
>    `head(interviews)`, keeping just the first through 6th rows of the interviews
>    dataset.
>
> > ## Solution
> >
> > ```{r}
> > ## 1.
> > interviews_100 <- interviews[100, ]
> > ## 2.
> > # Saving `n_rows` to improve readability and reduce duplication
> > n_rows <- nrow(interviews)
> > interviews_last <- interviews[n_rows, ]
> > ## 3.
> > interviews_middle <- interviews[(n_rows / 2), ]
> > ## 4.
> > interviews_head <- interviews[-(7:n_rows), ]
> > ```


