# Advanced variable creation with `forcats` {#forcats}

teaching: 60    
exercises: 15   
adapted from:  
questions:  

- "How can I easily create new categorical variables?" 

objectives:  

- 

keypoints:  

- 

## ForCats

## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered or unordered. While factors look (and often
behave) like character vectors, they are actually treated as integer vectors by
R. So you need to be very careful when treating them as strings.

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
respondent_floor_type <- factor(c("earth", "cement", "cement", "earth"))
```

R will assign `1` to the level `"cement"` and `2` to the level `"earth"`
(because `c` comes before `e`, even though the first element in this vector is
`"earth"`). You can see this by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(respondent_floor_type)
nlevels(respondent_floor_type)
```

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. Here, one way to reorder our levels in the `respondent_floor_type` vector would be:

```{r, results=TRUE, purl=FALSE}
respondent_floor_type # current order
respondent_floor_type <- factor(respondent_floor_type, levels = c("earth", "cement"))
respondent_floor_type # after re-ordering
```

In R's memory, these factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"cement"`,
`"earth"` is more descriptive than `1`, and `2`. Which one is "earth"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels. It also makes renaming levels easier. Let's say we made a mistake and need to recode "cement" to "brick".

```{r, results=TRUE, purl=FALSE}
levels(respondent_floor_type)
levels(respondent_floor_type)[2] <- "brick"
levels(respondent_floor_type)
respondent_floor_type
```



### Converting factors

If you need to convert a factor to a character vector, you use
`as.character(x)`.

```{r, purl=FALSE}
as.character(respondent_floor_type)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to numbers.
Another method is to use the `levels()` function. Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)                     # Wrong! And there is no warning...
as.numeric(as.character(year_fct))       # Works...
as.numeric(levels(year_fct))[year_fct]   # The recommended way.
```

Notice that in the recommended `levels()` approach, three important steps occur:

* We obtain all the factor levels using `levels(year_fct)`
* We convert these levels to numeric values using `as.numeric(levels(year_fct))`
* We then access these numeric values using the underlying integers of the
vector `year_fct` inside the square brackets

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's extract the `memb_assoc` column from our data frame, convert it into a
factor, and use it to look at the number of interview respondents who were or
were not members of an irrigation association:

```{r factor-plot-default-order, purl=TRUE}
## create a vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc
## convert it into a factor
memb_assoc <- as.factor(memb_assoc)
## let's see what it looks like
memb_assoc
## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```

Looking at the plot compared to the output of the vector, we can see that n
addition to "no"s and "yes"s, there are about some respondents for which the
information about whether they were part of an irrigation association hasn't
been recorded, and encoded as missing data. They do not appear on the plot.
Let's encode them differently so they can counted and visualized in our plot.


```{r factor-plot-reorder, purl=TRUE}
## Let's recreate the vector from the data frame column "memb_assoc"
memb_assoc <- interviews$memb_assoc
## replace the missing data with "undetermined"
memb_assoc[is.na(memb_assoc)] <- "undetermined"
## convert it into a factor
memb_assoc <- as.factor(memb_assoc)
## let's see what it looks like
memb_assoc
## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(memb_assoc)
```


> ## Exercise
>
> * Rename the levels of the factor to have the first letter in uppercase:
>   "No","Undetermined", and "Yes".
>
> * Now that we have renamed the factor level to "Undetermined", can you
>   recreate the barplot such that "Undetermined" is last (after "Yes")?
>
> > ## Solution
> >
> > ```{r factor-plot-exercise}
> > levels(memb_assoc) <- c("No", "Undetermined", "Yes")
> > memb_assoc <- factor(memb_assoc, levels = c("No", "Yes", "Undetermined"))
> > plot(memb_assoc)
> ```
> {: .solution}
{: .challenge}

