# Advanced variable creation with `forcats` {#forcats}

teaching: 60    
exercises: 15   
adapted from:  
questions:  

- "How can I easily create new categorical variables?" 

objectives:  

- 

keypoints:  

- 

```{r, eval=TRUE,  message=FALSE, purl=FALSE}
library(tidyverse)
interviews <- read_csv("data/SAFI_clean.csv", na = "NULL")
```

## Factors

R has a special data class, called factor, to deal with categorical data that
you may encounter when creating plots or doing statistical analyses. Factors are
very useful and actually contribute to making R particularly well suited to
working with data. So we are going to spend a little time introducing them.

Factors represent categorical data. They are stored as integers associated with
labels and they can be ordered or unordered. While factors look (and often
behave) like character vectors, they are actually treated as integer vectors by
R. So you need to be very careful when treating them as strings.

Factors are particularly useful when making plots or running statistical models.
Unfortunately, they can also be very tricky to work with, because they are secretly
numbers behind the scenes. Working with factors in base R can lead to errors that
are almost impossible for human analysts to catch, but there is a `tidyverse` package
that makes it much easier to work with factors, and prevents many common mistakes.
It is called `forcats` (an anagram of the word "factors" and also because it is a
package **for** working with **cat**egorical variable**s**). 

Let's load the `forcats` package so we can use the functions it comes with

```{r}
library(forcats)
```

Once created, factors can only contain a pre-defined set of values, known as
*levels*. By default, base R always sorts levels in alphabetical order. For
instance, if you have a factor with 2 levels:

```{r, purl=TRUE}
factor(c("earth", "cement", "cement", "earth"))
```

R will assign `1` to the level `"cement"` and `2` to the level `"earth"`
(because `c` comes before `e` in the alphabet, even though the first element in this vector is`"earth"`). 

In R's memory, factors are represented by integers (1, 2), but are more
informative than integers because factors are self describing: `"cement"`,
`"earth"` is more descriptive than `1`, and `2`. Which one is "earth"? You
wouldn't be able to tell just from the integer data. Factors, on the other hand,
have this information built in. It is particularly helpful when there are many
levels.

However, the default ordering of levels in base R is less than ideal, 
because it depends on the language you have set for your R session, and can 
lead to un-reproducble code. 

In the `forcats` package, there is a function that makes a factor but creates 
the levels in the order they appear. 

```{r purl=TRUE}
respondent_floor_type <- as_factor(c("earth", "cement", "cement", "earth"))
respondent_floor_type
```


You can see the levels and their order by using the function `levels()` and you can find
the number of levels using `nlevels()`:

```{r, purl=FALSE}
levels(respondent_floor_type)
nlevels(respondent_floor_type)
```

## reordering factor levels

Sometimes, the order of the factors does not matter, other times you might want
to specify the order because it is meaningful (e.g., "low", "medium", "high"),
it improves your visualization, or it is required by a particular type of
analysis. In `forcats`, one way to reorder our levels in the `respondent_floor_type` 
vector would be:

```{r, results=TRUE, purl=FALSE}
respondent_floor_type # current order
respondent_floor_type <- fct_relevel(respondent_floor_type, "cement", "earth")
respondent_floor_type # after re-ordering
```

This is perhaps easier to see with a few more factor levels. Let's use our real data,

```{r}
respondent_floor_type <- as_factor(interviews$respondent_wall_type)
levels(respondent_floor_type)
```

Say we want `sunbricks` to come first in the factor order. We can use `fct_relevel` 
to move it up.

```{r, results=TRUE, purl=FALSE}
respondent_floor_type # current order
respondent_floor_type <- fct_relevel(respondent_floor_type, "sunbricks")
respondent_floor_type # after re-ordering
```

The `fct_relevel` function allows you to move any number of levels to any location.
If you re-specify the entire list of levels, it will re-order the whole list. 
But, if you just specify one level (like we did here) that level gets moved to 
the front of the list. 

Most of the useful functions in `forcats` start with `fct_`, so you can try typing 
that much in your Console or code chunk and hitting Tab to use the code completion 
feature in RStudio to see what some of the other functions are called. 

Another way to re-order your factor levels is by frequency, so the most common 
factor levels come first, and the less common come later. (This is often useful
for plotting!)

```{r}
levels(respondent_floor_type)
respondent_floor_type <- fct_infreq(respondent_floor_type, ordered = TRUE)
levels(respondent_floor_type)
```

## renaming factor levels

`forcats` makes easy to rename factor levels. 
Let's say we made a mistake and need to recode "cement" to "brick". We'd use the 
`fct_recode` function to do this. 

```{r, results=TRUE, purl=FALSE}
levels(respondent_floor_type)

respondent_floor_type <- fct_recode(respondent_floor_type, brick = "cement")

levels(respondent_floor_type)
respondent_floor_type
```



### Converting factors

If you are working in the tidyverse, your data will come in with very few factor 
variables by default. So, the most common conversion you will need to do is from 
a character vector to a factor, like we did before,

```{r}
as_factor(c("earth", "cement", "cement", "earth"))
```


If you find yourself with many factors you need to convert to character vectors, 
it's good to check your read-in code first, to see if you accidentally used 
`read.csv` instead of `read_csv`. In the unlikely event you need to convert a 
factor to a character vector, you use `as.character(x)`.

```{r, purl=FALSE}
as.character(respondent_floor_type)
```

Converting factors where the levels appear as numbers (such as concentration
levels, or years) to a numeric vector is a little trickier. This is where many
mistakes with factor variables are made. The `as.numeric()`
function returns the index values of the factor, not its levels, so it will
result in an entirely new (and unwanted in this case) set of numbers.
One method to avoid this is to convert factors to characters, and then to numbers.
Compare:

```{r, purl=TRUE}
year_fct <- factor(c(1990, 1983, 1977, 1998, 1990))
as.numeric(year_fct)                     # Wrong! And there is no warning...
as.numeric(as.character(year_fct))       # The recommended way.
```

Again, if you find yourself doing this, you may want to check your data read-in 
code! 

### Renaming factors

When your data is stored as a factor, you can use the `plot()` function to get a
quick glance at the number of observations represented by each factor level.
Let's convert the `memb_assoc` column from our data frame into a
factor, and use it to look at the number of interview respondents who were or
were not members of an irrigation association:

```{r factor-plot-default-order, purl=TRUE}
## convert it into a factor
interviews <- interviews %>%
  mutate(memb_assoc = as_factor(memb_assoc))
interviews$memb_assoc

## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(interviews$memb_assoc)
```

Looking at the plot compared to the output of the vector, we can see that n
addition to "no"s and "yes"s, there are about some respondents for which the
information about whether they were part of an irrigation association hasn't
been recorded, and encoded as missing data. They do not appear on the plot.
Let's encode them differently so they can counted and visualized in our plot.


```{r factor-plot-reorder, purl=TRUE}
## replace the missing data with "undetermined"
interviews <- interviews %>%
  mutate(memb_assoc = fct_explicit_na(memb_assoc, na_level = "undetermined"))

interviews$memb_assoc
## bar plot of the number of interview respondents who were
## members of irrigation association:
plot(interviews$memb_assoc)
```

Again, if we wanted to reorder our factor here in order of 

```{r}
interviews <- interviews %>%
  mutate(memb_assoc = fct_infreq(memb_assoc, ordered = TRUE))
plot(interviews$memb_assoc)


interviews <- interviews %>%
  mutate(memb_assoc = fct_relevel(memb_assoc, "No", "Yes", "Undetermined"))
```


> ## Exercise
>
> * Rename the levels of the factor to have the first letter in uppercase:
>   "No","Undetermined", and "Yes".
>
> * Now that we have renamed the factor level to "Undetermined", can you
>   recreate the barplot such that "Undetermined" is last (after "Yes")?
>
> > ## Solution
> >
> > ```{r factor-plot-exercise}
> > interviews <- interviews %>%
> >  mutate(memb_assoc = fct_recode(memb_assoc, No = "no", Yes = "yes", 
> >                                 Undetermined = "undetermined"))
> > plot(interviews$memb_assoc)
> > interviews <- interviews %>%
  mutate(memb_assoc = fct_relevel(memb_assoc, "No", "Yes", "Undetermined"))
> > plot(interviews$memb_assoc)
> ```
> {: .solution}
{: .challenge}

